To Batch or Not to Batch: Test-Ordering Variability in the ED

### 2024-09-09 manuscript/manuscript.tex
getting going

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}

@@

Our research design focuses on adults who visit the Mayo Clinic of
Arizona ED. We observe approximately \(48,000\) such visits during the
study period. To improve power, we drop encounters with rare chief
complaints (\(<1000\) total encounters of this kind) and complaints
where a batch order occurs less than \(10\) percent of the time. Since
batch orders are rare for these cases, our physician batch tendency
instrument could suffer from a weak instrument problem if we included
them. Examples of complaints dropped include urinary complaints and
allergic reactions. Excluding these conditions does not introduce
selection bias unless physician test batching tendency is orthogonal to
physician diagnosing behavior. While this assumption may be violated if
we were to use a very detailed level of chief complaint information upon
which to base our exclusion criterion, it is plausibly satisfied when
using broad complaint
categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
In order to estimate a precise measure of physician-level batch
tendency, we further restrict our sample to the \(42,000\) encounters
involving physicians who treat over 500 ED cases per year.

@@

Our explanatory variable in the IV analysis, \(Batched_i\), is an
indicator for whether patient \(i\) has their tests batch-ordered at
their ED encounter. While the patient could decide not to undergo the
tests ordered by the physician, this is rare in practice. Below, we
detail our primary outcomes: (a) ED length of stay (LOS), (b) resource
utilization, and (c) 72-hour return.

@@

\emph{(c) 72-hour return}-- The rate of patients returning to the ED
within 72 hours of their initial visit is a proxy for the quality of
care received. A high 72-hour return rate might indicate inadequate
treatment or diagnostic oversight during the initial visit. We measure
this outcome as the proportion of patients with a subsequent ED visit
within 72 hours post-initial discharge. The hypothesis is that batch
testing may lead to variations in the 72-hour return rate, potentially
influencing the quality of care received.

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure
\ref{fig:physician_batching} graphs the histogram of batch-ordering
frequency by physician for popular chief complaints, highlighting that
the variation in batching differs systematically.
\hyperref[tab:regression]{Table 3} presents the ``first stage'' in a
regression table: being assigned to a \(10\)pp higher batch-tendency
physician is associated with a \(8\)pp increase in the likelihood of
having tests batch-ordered in the ED. The F-statistic is 50 when all
controls and fixed effects are included. The coefficient is greater than
one because all emergency visits are used to construct the tendency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rare complaints.

\begin{figure}[h]
  \centering
  \caption{Batch Frequency by Physician for High Volume Complaints}
  \label{fig:physician_batching}
  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
\begin{tablenotes}
\tiny
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests. Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue). The quartile of batchers is calculated based on the batching rate across all complaint areas per physician. Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching. Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
\end{tablenotes}  
\end{figure}

\begin{table}[ht]
\centering
\caption{First Stage: Effect of Batch Tendency on Test Batching}
\label{tab:regression}
\begin{tabular}{p{9cm}ccc}
\toprule
\textbf{} & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 1.06 *** & 1.08 *** & 1.08 *** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43327 & 43327 & 43327 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

In this section, we explore the causal influence of physician batch
tendency on patient outcomes and resource utilization in the emergency
department. Our reduced form regression results, as presented in Model 1
of Table \ref{tab:reducedform}, reveal that batch tendency correlates
with prolonged ED length of stay (LOS), increased test ordering, and
does not materially alter the likelihood of a 72-hour return. Notably,
these associations are nuanced by an underlying factor: the physician's
testing inclination.

\begin{table}[htbp]
\centering
\caption{Reduced-Form Results: Length of Stay, Number of Tests, and 72-Hour Return}
\label{tab:reducedform}
\begin{tabular}{p{6.5cm}ccc}
\toprule
& \multicolumn{1}{c}{ln(LOS)}
& \multicolumn{1}{c}{Number of Tests}
& \multicolumn{1}{c}{72-Hour Return} \\
& \multicolumn{1}{c}{(1)}
& \multicolumn{1}{c}{(2)}
& \multicolumn{1}{c}{(3)} \\
\midrule
\textit{Model 1} & & & \\
Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
\midrule
\textit{Model 2} & & & \\
Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
                    &     &     &  \\
Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Estimates of the reduced form are for the baseline sample described in the text. Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

We construct a physician's testing inclination using the same method
used to measure physician batch tendency, but instead we use the
physician's residualized leave-out average test rate to get a measure of
a physicians propensity to test. A key consideration in our analysis is
the strong positive correlation (\(r = 0.79\)) between batch tendency
and testing inclination. Such a high degree of correlation raises the
concern that physicians with a higher testing inclination may batch more
often as a result of ordering more tests. Failure to control for this
inclination could result in biased estimates of the effect of batch
tendency.

The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG})
elucidates the interconnected pathways between batch tendency, batch
ordering, and primary outcomes. We posit that while batch tendency
directly influences the practice of batch ordering tests, batch tendency
and batch ordering are concurrently influenced by testing inclination.
This is critical as testing inclination has a direct effect on primary
outcomes, necessitating its inclusion as a control variable to
disentangle its confounding effects.

\begin{figure}[h]
  \centering
  \caption{Directed Acyclic Graph}
  \label{fig:DAG}
  \includegraphics[width=4in]{../outputs/figures/DAG.png}
\begin{tablenotes}
\small
\item \textit{}
\end{tablenotes}  
\end{figure}

In Model 2 of Table \ref{tab:reducedform}, we control for testing
inclination and observe notable changes in the coefficients for batch
tendency. The previously significant positive association with ln(LOS)
diminishes, suggesting that the initial estimate may have been capturing
the combined effects of batching and testing inclination. Estimated
coefficients for batch tendency are scaled by the difference in batch
tendency going from the ninetieth to tenth percentile in physician batch
tendency for interpretability, and coefficients for testing inclination
are scaled by the difference in testing inclination going from the
ninetieth to tenth percentile in physician testing inclination.
Assignment to a physician in the top batching decile (relative to one in
the bottom decile) is associated with a 0.2 unit increase in the number
of diagnostic tests being ordered per patient encounter, without any
additional change in the likelihood of a 72-hour return.

The effects of testing inclination are also striking. We find that a
movement from a physician in the top testing decile (relative to one in
the bottom decile) is associated with 0.6\% change in the length of stay
in the emergency department, a 9.6 excess diagnostic tests being ordered
per 100 patient encounters, and -0.03 pp change in the probability of a
patient returning within 72 hours.

@@

In this section we examine the causal effects of batch ordering in the
ED. Mirroring our presentation of the reduced-form results, we turn to
presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED
LOS, number of tests, and 72-hour return.

\begin{table}[!htbp] \centering 
  \caption{2SLS Results: Length of Stay, Number of Tests, and 72-Hour Return} 
  \label{tab:2SLS} 
\begin{tabular}{p{7cm}ccc}
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & ln(LOS) & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Batch & $-$0.086 & 0.135$^{***}$ & 0.018 \\ 
  & (0.419) & (0.046) & (0.018) \\ 
  & & & \\ 
 Testing Inclination & 0.210$^{**}$ & 0.335$^{***}$ & $-$0.009$^{***}$ \\ 
  & (0.082) & (0.008) & (0.003) \\ 
  & & & \\ 
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.199 & 0.312 & 0.008 \\ 
Adjusted R$^{2}$ & 0.194 & 0.307 & 0.002 \\ 
Residual Std. Error (df = 41667) & 0.496 & 0.817 & 0.184 \\ 
\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Estimates of the IV model are for the baseline sample described in the text. Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

\emph{ED length of stay (LOS).}-- Our analysis reveals that batch
ordering does not significantly reduce the length of stay in the ED.
This finding holds even after accounting for various control variables,
including ED volume and patient acuity. Contrary to common assumptions
that batching tests could expedite patient throughput, our results
suggest that the efficiency gains could be offset. A possible
explanation lies in excess tests being ordered due to batching, which
could lead to increased wait times for patients. This is consistent with
our finding that batching increases the number of tests ordered.

\emph{Resource Utilization.}-- The impact of batching on resource
utilization is pronounced. We observe a 0.135 increase in the number of
tests ordered due to batching, corresponding to approximately 2,206
additional tests during our study period. This is likely due to the
information gain that occurs from sequential test ordering, where a
physician may order additional tests after receiving the results of the
first test.

\emph{72-hour Return}-- Our findings indicate that batch ordering does
not significantly impact the likelihood of a 72-hour return. This
suggests that despite the increased number of tests, there is no
corresponding decrease in short-term readmission rates. This highlights
the need for more targeted testing strategies that balance comprehensive
assessment with resource efficiency.

The instrumental variable analysis presented here underscores the
complexities underlying batch ordering practices in the ED. While
batching is often seen by physicians as a method to enhance efficiency,
our findings suggest that it may lead to increased resource utilization
without corresponding gains in patient throughput or reductions in
short-term return rates. These insights point to the need for a more
nuanced approach to test ordering in emergency care, one that carefully
weighs the benefits of comprehensive assessment against the risks and
costs of potential over-testing.

@@

\begin{tabular}{@{\extracolsep{10pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textbf{1-minute window}} \\
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Testing Inclination & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
  & (0.203) & (0.018) & (0.009) \\ 
  & & & \\ 
 Batch & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
  & (0.496) & (0.054) & (0.023) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
Residual Std. Error & 0.492 & 0.836 & 0.184 \\ 
\hline 
\\[-1.8ex] & \multicolumn{3}{c}{\textbf{10-minute window}} \\
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Testing Inclination & 0.650$^{***}$ & 0.957$^{***}$ & $-$0.027$^{**}$ \\ 
  & (0.241) & (0.022) & (0.011) \\ 
  & & & \\ 
 Batch & $-$0.205 & 0.141$^{***}$ & 0.017 \\ 
  & (0.438) & (0.045) & (0.018) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.146 & 0.315 & 0.009 \\ 
Adjusted R$^{2}$ & 0.141 & 0.311 & 0.003 \\ 
Residual Std. Error & 0.512 & 0.814 & 0.184 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular}
\begin{tablenotes}
\tiny
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

>>>>

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}
\usepackage{caption}

@@

Our research design focuses on adults who visit the Mayo Clinic of
Arizona ED. We observe approximately \(48,000\) such visits during the
study period. To improve power, we drop encounters with rare chief
complaints (\(<1000\) total encounters of this kind) and complaints
where a batch order occurs less than \(10\) percent of the time. Since
batch orders are rare for these cases, our physician batch tendency
instrument could suffer from a weak instrument problem if we included
them. Examples of complaints dropped include urinary complaints and
allergic reactions. Excluding these conditions does not introduce
selection bias unless physician test batching tendency is orthogonal to
physician diagnosing behavior. While this assumption may be violated if
we were to use a very detailed level of chief complaint information upon
which to base our exclusion criterion, it is plausibly satisfied when
using broad complaint categories. In order to estimate a precise measure
of physician-level batch tendency, we further restrict our sample to the
\(42,000\) encounters involving full-time physicians who treat over 500
ED cases per year.

@@

Our explanatory variable in the IV analysis, \(Batched_i\), is an
indicator for whether patient \(i\) has their tests batch-ordered at
their ED encounter. While the patient could decide not to undergo the
tests ordered by the physician, this is rare in practice. Below, we
detail our primary outcomes: (a) ED length of stay (LOS), and (b)
resource utilization.

@@

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure
\ref{fig:physician_batching} graphs the histogram of batch-ordering
frequency by physician for popular chief complaints, highlighting that
the variation in batching differs systematically.
\hyperref[tab:regression]{Table 3} presents the ``first stage'' in a
regression table: being assigned to a \(10\)pp higher batch-tendency
physician is associated with a \(6\)pp increase in the likelihood of
having tests batch-ordered in the ED. The F-statistic is 94 when all
controls and fixed effects are included. The coefficient is greater than
one because all emergency visits are used to construct the tendency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rare complaints.

\begin{figure}[h]
  \centering
  \caption{Batch Frequency by Physician for High Volume Complaints}
  \label{fig:physician_batching}
  \includegraphics[width=1\textwidth]{../outputs/figures/batch_ordering.png}
\begin{tablenotes}
\small
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests. Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue). The quartile of batchers is calculated based on the batching rate across all complaint areas per physician. Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching. Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
\end{tablenotes}  
\end{figure}

\begin{table}[ht]
\centering
\caption{First Stage: Effect of Batch Tendency on Test Batching}
\label{tab:regression}
\begin{tabular}{p{9cm}ccc}
\toprule
\textbf{} & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 1.06 *** & 1.06 *** & 1.06 *** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model) & 23.87 & 94.41 & 90.44  \\  
F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 41929 & 41929 & 41929 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

In this section, we explore the causal influence of physician batch
tendency on patient outcomes and resource utilization in the emergency
department. We posit that while batch tendency directly influences the
practice of batch ordering tests, both batch tendency and batch ordering
are concurrently influenced by a physician's testing inclination. Given
that testing inclination directly affects primary outcomes, we include
it as a control variable in our regression models to mitigate its
confounding effects.

To quantify a physician's testing inclination, we employ a similar
approach to that used in measuring physician batch tendency.
Specifically, we calculate the physician's residualized leave-out
average test rate, which serves as a proxy for their propensity to order
tests. It is important to note the strong positive correlation
(\(r = 0.79\)) between batch tendency and testing inclination. This
substantial correlation suggests that physicians with a higher
propensity to test may also exhibit a higher tendency to batch orders,
potentially as a consequence of their testing strategies. Neglecting to
account for testing inclination could lead to overestimated effects of
batch tendency due to omitted variable bias.

Reduced-form regression results in Table \ref{tab:reducedform} reveal
the effect the average total effect of batch tendency on resource
utilization and LOS. However, it is important to note that the estimated
total effect of batch tendency accounts for both direct and indirect
(mediated) effects of batch tendency. We hypothesize that while batch
ordering may streamline the testing process, resulting in a quicker
completion of a given number of tests, it simultaneously appears to lead
to an increase in the total number of tests ordered. The increased
testing volume, in turn, is associated with an extended LOS.

\begin{table}[!htbp] \centering 
\caption{Reduced-Form Results: Length of Stay and Number of Tests}
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & ln(LOS) & Number of Tests \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
 Batch Tendency & $-$0.109 & 0.149$^{***}$ \\ 
  & (0.464) & (0.048) \\ 
  & & \\ 
 Testing Inclination & 0.605$^{**}$ & 0.952$^{***}$ \\ 
  & (0.240) & (0.023) \\ 
  & & \\ 
Seasonality and shift fixed effects? & Yes & Yes \\
Chief Complaint? & Yes & Yes  \\
\hline 
\textit{N} & 41,929 & 41,929 \\ 
R$^{2}$ & 0.215 & 0.265 \\ 
Adjusted R$^{2}$ & 0.213 & 0.263 \\ 
Residual Std. Error (df = 41800) & 0.490 & 0.842 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\small
\item \textit{Estimates of the reduced form are for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

We introduce a Directed Acyclic Graph (DAG) to illustrate the
interconnected pathways between batch tendency, batch ordering, and
primary outcomes. The DAG in Figure \ref{fig:DAG} shows that batch
tendency directly influences batch ordering, which in turn affects the
number of tests ordered. The number of tests ordered, in turn,
influences the length of stay. The DAG also illustrates the effect of
batch tendency on the length of stay is mediated by the number of tests
ordered.

\begin{figure}[h]
  \centering
  \caption{Directed Acyclic Graph}
  \label{fig:DAG}
  \includegraphics[width=6in]{../outputs/figures/dagitty.png}
\begin{tablenotes}
\small
\item \textit{}
\end{tablenotes}  
\end{figure}

To determine whether or not the effect of batch tendency on LOS is
mediated by the number of tests ordered, we estimate a mediation
analysis using the approach outlined by \citet{imai2010general}. Our
analysis, scrutinizes the pathways through which physician batch
tendency influences patient length of stay (LOS) in the emergency
department. The analysis delineates both direct effects, which capture
the influence of batch tendency on LOS without intermediaries, and
indirect effects, which operate through the mediator of test ordering
volume. Our findings underscore a significant Average Direct Effect
(ADE) of batch tendency on LOS, estimated at -0.15 (p = 0.046).

@@

In this section we examine the causal effects of batch ordering in the
ED. Mirroring our presentation of the reduced-form results, we turn to
presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED
LOS and number of tests ordered. We first start by estimating the
effects of batch ordering on the number of tests ordered using the
standard 2SLS approach. Next we estimate the direct and indirect effects
of batch ordering on LOS using a causal mediation analysis.

\begin{table}[!htbp]
\centering
\begin{threeparttable}
\caption{Combined 2SLS and Causal Mediation Analysis Results}
\label{tab:combined_results}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
\toprule
& \multicolumn{2}{c}{2SLS Results} \\
\cmidrule{2-3}
& ln(LOS) & Number of Tests \\
\midrule
Batch & -0.086 & 0.135$^{***}$ \\
& (0.419) & (0.046) \\
& & \\
\addlinespace
Seasonality and Shift Fixed Effects & Yes & Yes \\
Chief Complaint and Severity & Yes & Yes \\
\midrule
Observations & 41,901 & 41,901 \\
R$^{2}$ & 0.183 & 0.310 \\
\bottomrule
\end{tabular*}

\smallskip % Adds some space before the next part of the table

\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
\toprule
& \multicolumn{2}{c}{Causal Mediation Analysis} \\
\cmidrule{2-3}
Effect & Estimate & 95\% CI \\
\midrule
ACME & 0.036 & [-0.009, 0.08] \\
ADE & -0.136 & [-0.221, -0.04] \\
Total Effect & -0.100 & [-0.195, 0.00] \\
Proportion Mediated & -0.333 & [-2.831, 1.21] \\
\bottomrule
\end{tabular*}

\begin{tablenotes}
\small
\item The 2SLS results are based on robust standard errors clustered at the physician level, with controls for seasonality, shift fixed effects, chief complaint, and severity. The causal mediation analysis examines whether the number of tests mediates the effect of batching on length of stay, using a quasi-Bayesian confidence interval approach with 1000 simulations.
\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
\end{tablenotes}
\end{threeparttable}
\end{table}

\emph{ED length of stay (LOS).}-- Our causal mediation analysis reveals
that the Average Direct Effect (ADE) of test batching on the natural log
of LOS is -0.13607, indicating that, when not considering the mediating
effect of the number of tests, batching leads to a decrease in the
expected LOS. Specifically, this translates to an approximate 12.7
percent reduction in LOS when tests are batched, all else being equal.
This result, complemented by our 2SLS estimate of -0.086, suggests that
the direct effect of batch ordering on LOS is negative. Our results also
suggest a ``washing out'' effect, where the effeciency gains from
batching are offset by the increased number of tests ordered. This is
consistent with our finding that batching increases the number of tests
ordered.

\emph{Resource Utilization.}-- The impact of batching on resource
utilization is pronounced. We observe a 0.135 increase in the number of
tests ordered due to batching, corresponding to approximately 13.5
excess tests per 100 batch orders, where excess means tests that would
not have been ordered had the physician not batched. This result is
consistent with the notion that batching leads to increased resource
utilization in the ED.

The instrumental variable analysis presented here underscores the
complexities underlying batch ordering practices in the ED. While
batching is often seen by physicians as a method to enhance efficiency,
our findings suggest that it may lead to increased resource utilization
and that the corresponding gains in patient throughput are washed out.
These insights point to the need for a more nuanced approach to test
ordering in emergency care, one that carefully weighs the benefits of
comprehensive assessment against the risks and costs of potential
over-testing.

@@

\begin{tabular}{@{\extracolsep{10pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textbf{1-minute window}} \\
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Testing Inclination & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
  & (0.203) & (0.018) & (0.009) \\ 
  & & & \\ 
 Batch & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
  & (0.496) & (0.054) & (0.023) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
Residual Std. Error & 0.492 & 0.836 & 0.184 \\ 
\hline 
\\[-1.8ex] & \multicolumn{3}{c}{\textbf{10-minute window}} \\
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Testing Inclination & 0.650$^{***}$ & 0.957$^{***}$ & $-$0.027$^{**}$ \\ 
  & (0.241) & (0.022) & (0.011) \\ 
  & & & \\ 
 Batch & $-$0.205 & 0.141$^{***}$ & 0.017 \\ 
  & (0.438) & (0.045) & (0.018) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.146 & 0.315 & 0.009 \\ 
Adjusted R$^{2}$ & 0.141 & 0.311 & 0.003 \\ 
Residual Std. Error & 0.512 & 0.814 & 0.184 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular}
\begin{tablenotes}
\small
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{2SLS Results for the "Batch" Variable Across Different General Complaints: Outcome of Number of Tests}
\begin{tabular}{lcc}
\hline
\textbf{Complaint} & \textbf{Batching Coefficient} & \textbf{Standard Error} \\
\hline
Musculoskeletal/Extremity & 0.476** & (0.235)  \\
Respiratory-Related & 0.244 & (0.209)   \\
General/Other Symptoms & 0.179 & (0.207)   \\
Gastrointestinal/Abdominal & -0.092 & (0.128)   \\
Cardiac/Chest-Related & 0.270 & (0.264)   \\
Neurological/Syncope & -0.080 & (0.493)   \\
\hline
\end{tabular}
\label{tab:batch_variable_outcome}
\end{table}
